Categories of Optics

### 2018-06-08 paper.tex
Long missing proof of adjointness

@@

\todo{For \emph{lawful} affine traversals, there may not be a difference}

@@

\subsection{Polymorphic Optics}
Haskell's optics \todo{find an older reference from DB theory?} allow \emph{polymorphic updates}, where the type of the codomain of the lens can be changed by an update, causing a corresponding change in the type of the domain. As an example, the lens into the first entry of a tuple has type
\begin{center}
\texttt{first :: Lens (a, b) (a', b) a a'}, 
\end{center}
then:
\begin{center}
\texttt{set first (1, 5) "hello" == ("hello", 5)}
\end{center}
Note that the type has been changed from $\texttt{(Int, Int)}$ to $\texttt{(String, Int)}$.

@@

\subsection{Tambara Modules}
Let $I = \C(-,{=}) : \C \hto \C$ be the identity profunctor and $\odot$ be profunctor composition, written in diagrammatic order. The following section generalises definitions that first appeared in \cite{Doubles}.
\begin{definition}
Suppose $\C$ is acted on by $(\M, \otimes, I)$ and let $P \in \Prof(\C, \C)$ be a profunctor. A \emph{Tambara module structure for $\M$} on $P$ is a family of maps:
\begin{align*}
\alpha_{A,B,M} : P(A,B) \to P(MA, MB)
\end{align*}
natural in $A$ and $B$, and dinatural in $M$. Further, we must have that $\alpha_{A,B,I} = \id_{P(A,B)}$, and that the following diagram commutes:
\[
\begin{tikzcd}
P(A,B) \ar[r, "\alpha_{A,B,M}"] \ar[dr, "\alpha_{A, B, N \otimes M}" below left] & P(MA, MB) \ar[d, "\alpha_{MA, MB, N}" right] \\
& P(NMA, NMB)
\end{tikzcd}
\]
for all $A, B \in \C$ and $N, M \in \M$.
\end{definition}

Note that the hom profunctor $I := \C(-, =) : \C^\op \times \C \to \Set$ has a canonical Tambara module structure for any $\M$, given by functoriality.

@@

\begin{definition}
There is a category $\Tamb(\M)$ of Tambara modules and natural transformations that respect the extra structure, in the sense that for any $l : P \to Q$, the diagram
\[
\begin{tikzcd}
P(A,B) \ar[r, "\alpha_{A,B,M}"] \ar[d, "l_{A,B}" left] & P(MA, MB) \ar[d, "l_{MA, MB}" right] \\
Q(A,B) \ar[r, "\beta_{A,B,M}"] & Q(MA, MB)
\end{tikzcd}
\]
commutes.
\end{definition}

This category is monoidal with respect to $\odot$ as given above, and the monoidal unit is given by $I$. There is an evident forgetful functor $U : \Tamb(\M) \to \Prof(\C, \C)$ that is strong monoidal. This forgetful functor has both a left and right adjoint; particularly important for us is the left adjoint:

\begin{definition}
Let $\Pastro_\M : \Prof(\C, \C) \to \Tamb(\M)$ be the functor:
\begin{align*}
\Pastro_\M(P) := \int^{M \in \M}  \C(-, M{=}) \odot P \odot \C(M-, {=}) 
\end{align*}
Or, in other words, 
\begin{align*}
\mathrm{Pastro}_\C(\M)(P)(A,B) := \int^{M \in \M} \int^{C,D \in \C} \C(A, MC) \times P(C,D) \times  \C(MD, B)
\end{align*}
The module structure $\alpha_{A,B,M} : \Pastro_\M(P)(A,B) \to \Pastro_\M(P)(MA, MB)
$ is given by map induced by 
\begin{align*}
&\int^{C,D \in \C} \C(A, NC) \times P(C,D) \times  \C(ND, B) \\
\xrightarrow{\text{functoriality}} \quad& \int^{C,D \in \C} \C(MA, MNC) \times P(C,D) \times  \C(MND, MB) \\
\xrightarrow{\copr_{MN}} \quad&\int^{N \in \M} \int^{C,D \in \C} \C(MA, NC) \times P(C,D) \times  \C(ND, MB) \\
= \quad&\Pastro_\M(P)(MA, MB)
\end{align*}
for all $N \in \M$.
\end{definition}

\begin{proposition}
$\Pastro_\M : \Prof(\C, \C) \to \Tamb(\M)$ is left adjoint to $U : \Tamb(\M) \to \Prof(\C, \C)$.
\end{proposition}
\begin{proof}
\todo{Could crunch through this manually, may be a neat way to do it using techniques from P/S}
\end{proof}

\begin{corollary}
$\Pastro_\M$ is oplax monoidal.
\end{corollary}
\begin{proof}
This follows abstractly as the left adjoint of a strong monoidal functor \cite{Kelly1974}.
\end{proof}

>>>>

\todo{Do laws? For \emph{lawful} affine traversals, there may not be a difference}

@@

@@

\subsection{Polymorphic Optics}
Haskell's optics \todo{find an older reference from DB theory?} allow \emph{polymorphic updates}, where the type of the codomain of the lens can be changed by an update, causing a corresponding change in the type of the domain. As an example, the lens into the first entry of a tuple has type
\begin{center}
\texttt{first :: Lens (a, b) (a', b) a a'},
\end{center}
\todo{Why is the spacing strange above?} then:
\begin{center}
\texttt{set first (1, 5) "hello" == ("hello", 5)}
\end{center}
Note that the type has been changed from $\texttt{(Int, Int)}$ to $\texttt{(String, Int)}$.

@@

\subsection{Tambara Modules}
Let $I = \C(-,{=}) : \C \hto \C$ be the identity profunctor and $\odot$ be profunctor composition, written in diagrammatic order. The following section generalises definitions that first appeared in \cite[Section 3]{Doubles} for monoidal categories to the more general case of a monoidal action.

\begin{definition}
Suppose $\C$ is acted on by $(\M, \otimes, I)$ and let $P \in \Prof(\C, \C)$ be a profunctor. A \emph{Tambara module structure for $\M$} on $P$ is a family of maps:
\begin{align*}
\alpha_{A,B,M} : P(A,B) \to P(MA, MB)
\end{align*}
natural in $A$ and $B$, and dinatural in $M$. Further, we must have that $\alpha_{A,B,I} = \id_{P(A,B)}$, and that the following diagram commutes:
\[
\begin{tikzcd}
P(A,B) \ar[r, "\alpha_{A,B,M}"] \ar[dr, "\alpha_{A, B, N \otimes M}" below left] & P(MA, MB) \ar[d, "\alpha_{MA, MB, N}" right] \\
& P(NMA, NMB)
\end{tikzcd}
\]
for all $A, B \in \C$ and $N, M \in \M$.
\end{definition}

Note that the identity profunctor $I$ has a canonical Tambara module structure for any $\M$ given by functoriality.

@@

\begin{definition}
There is a category $\Tamb(\M)$ of Tambara modules and natural transformations that respect the extra structure, in the sense that for any $l : P \hto Q$, the diagram
\[
\begin{tikzcd}
P(A,B) \ar[r, "\alpha_{A,B,M}"] \ar[d, "l_{A,B}" left] & P(MA, MB) \ar[d, "l_{MA, MB}" right] \\
Q(A,B) \ar[r, "\beta_{A,B,M}"] & Q(MA, MB)
\end{tikzcd}
\]
commutes.
\end{definition}

This category is monoidal with respect to $\odot$ as given above with monoidal unit $I$. There is an evident forgetful functor $U : \Tamb(\M) \to \Prof(\C, \C)$ that is strong monoidal. This forgetful functor has both a left and right adjoint; important for us is the left adjoint:

\begin{definition}[{\cite[Section 5]{Doubles}}]
Let $\Pastro_\M : \Prof(\C, \C) \to \Tamb(\M)$ be the functor:
\begin{align*}
\Pastro_\M(P) := \int^{M \in \M}  \C(-, M{=}) \odot P \odot \C(M-, {=}) 
\end{align*}
Or, in other words, 
\begin{align*}
\Pastro_\M(P)(A,B) := \int^{M \in \M} \int^{C,D \in \C} \C(A, MC) \times P(C,D) \times  \C(MD, B)
\end{align*}
The module structure $\alpha_{A,B,M} : \Pastro_\M P(A,B) \to \Pastro_\M P (MA, MB)
$ is induced by the maps
\begin{align*}
&\C(A, NC) \times P(C,D) \times  \C(ND, B) \\
\xrightarrow{\text{functoriality}} \quad& \C(MA, MNC) \times P(C,D) \times  \C(MND, MB) \\
\xrightarrow{\copr_{MN}} \quad&\int^{N \in \M} \int^{C,D \in \C} \C(MA, NC) \times P(C,D) \times  \C(ND, MB) \\
= \quad&\Pastro_\M P (MA, MB)
\end{align*}
for all $C, D \in \C$ and $N \in \M$. Equationally, this is $\alpha_{A,B,M}(\langle l, p, r \rangle) = \langle Ml, p, Mr \rangle$.
\end{definition}

\begin{proposition}
$\Pastro_\M : \Prof(\C, \C) \to \Tamb(\M)$ is left adjoint to $U : \Tamb(\M) \to \Prof(\C, \C)$.
\end{proposition}
\begin{proof}
For any $P \in \Prof(\C, \C)$, there is a map $\eta : P \hto U \Phi_\M P$, given by $\eta(p) = \langle \id_A, p, \id_B \rangle$.

Suppose we have an element $\langle l, p, r \rangle \in \Pastro_\M P(A,B)$. Then one can check that this element is equal to 
\begin{align*}
\langle l, p, r \rangle = \Pastro_\M P(l, r) \alpha_{A, B, M} \; \eta(p)
\end{align*}
where $\alpha_{A, B, M}$ is the module structure map for $\Pastro_\M P$.

Now suppose $T \in \Tamb(\M)$ is a Tambara module with structure map $\beta$. We would like to show that for any map $f : P \hto UT$, there exists a unique $\hat f : \Phi_\M P \hto T$ so that $f$ factors as \[P \xrightarrow{\eta} U \Phi_\M P \xrightarrow{U\hat f} UT \].

The data of such a map $\hat f : \Phi_\M P \hto T$ is a natural transformation between the underlying profunctors. For the factorisation property to hold, we must have that $\hat{f}\eta(p) = f(p)$ for any $p \in P(A,B)$, but then the action on the remainder of $\Phi_\M P(A, B)$ is fixed:
\begin{align*}
\hat{f}(\langle l, p, r) \rangle &= \hat{f}(\Pastro_\M P(l, r) \alpha_{A, B, M} \; \eta(p)) \\
&=T(l, r) \; \beta_{A, B, N} \; f(p)
\end{align*}
This establishes uniqueness. It remains to show that $\hat{f}$ so defined is actually a Tambara module morphism, but this is easy:
\begin{align*}
\hat{f}\alpha_{A,B,N}(\langle l, p, r \rangle)
&= \hat{f}(\langle Nl, p, Nr \rangle) \\
&= T(Nl, Nr) \; \beta_{A, B, NM} \; f(p) \\
&= T(Nl, Nr) \; \beta_{A, B, NM} \; f(p) \\
&= T(Nl, Nr) \beta_{MA,MB,N} \; \beta_{A, B, M} \; f(p) \\
&= \beta_{A,B,N} T(l, r) \; \beta_{A, B, M} \; f(p) \\
&= \beta_{A,B,N} \hat{f} (\langle l, p, r \rangle)
\end{align*}
\end{proof}

\begin{corollary}
$\Pastro_\M$ is oplax monoidal.
\end{corollary}
\begin{proof}
This follows from abstract nonsense as the left adjoint of a strong monoidal functor \cite{Kelly1974}.
\end{proof}
